---
title: 'BirdCLEF 2023 via `voice`'
author: 'Filipe J. Zabala'
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output:
  rmarkdown::html_vignette:
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
ini0 <- Sys.time()
```

## 0. Context
This is a vignette presenting the [`voice`](https://cran.r-project.org/package=voice) approach to [Kaggle BirdCLEF 2023](https://www.kaggle.com/competitions/birdclef-2023/) competition.


## 1. Libraries and functions

### 1.1 Installing
[`birdclef2023`](https://github.com/filipezabala/birdclef2023) contains summarized data from [Kaggle BirdCLEF 2023](https://www.kaggle.com/competitions/birdclef-2023/) competition audios. The summarization was made using [`voice`](https://github.com/filipezabala/voice), an R library with functions to easily deal with audio.

```{r, eval=FALSE}
# install packs
install.packages('voice', dep = TRUE)
devtools::install_github('filipezabala/birdclef2023')
```

### 1.2 Calling
The libraries are called, `geti` and `gaussian` functions are defined.

```{r, message=FALSE, warning=FALSE}
# packs
library(voice)
library(tidyverse)
library(birdclef2023)

# geti function
geti <- function(x,i){x[i]}

# gaussian function
gaussian <- function(x, mu = 0, sigma = 1){
  exp(-((x-mu)/sigma)^2)
}
```

## 2. Data
In this section is presented how to download, extract and summarize the bird's audio features. As the results are available at [`birdclef2023`](https://github.com/filipezabala/birdclef2023), it may be ommited in a first reading.

### 2.1 Downloading
The data may be downloaded by the [official Kaggle API]( https://github.com/Kaggle/kaggle-api). The zip file occupies around 5.3GB on disk. 

```{r, eval=FALSE}
url <- 'https://storage.googleapis.com/kaggle-competitions-data/kaggle-v2/44224/5188730/bundle/archive.zip?GoogleAccessId=web-data@kaggle-161607.iam.gserviceaccount.com&Expires=1685615147&Signature=QeSfA1%2Bumjv4ND27r%2F7ln3XzHk%2F52GDgNq3EDtphEAKxQUkypPQHaIVCxlLFAnhfqtTM3V4mny0KtttbCXgseNaweq%2Bx1f1TjtY6DEjP%2FksG%2B%2BX0TiaLgl06xPhh1SZ%2FiZcCaEtESL3SKAZ9Nq5%2FbSoNKTzghgPP2ET4ncUxc5UeaqR6%2BVTtSWLDCEe%2FEzPHOT64gaF9w4gAnqZjkl7GFthY034mXaBSULWl1Ul6K55vX%2FJfXbaxDzrJbVB1HlU0eax3MZcnyy9msHSwWEsH8N08pSYtWJZA4GmPoGUn5YrNRr5%2Bc7YYVKWKKPxmH%2FG7%2Fw7Y3gj7vd7etLLBzcjRBA%3D%3D&response-content-disposition=attachment%3B+filename%3Dbirdclef-2023.zip'
options(timeout = 9999)
download.file(url = url, '~/Downloads/birdclef-2023.zip', mode = 'wd')
```

### 2.2 Extracting
Taking a look.
```{r}
head(unzip('~/Downloads/birdclef-2023.zip', list = TRUE))
nrow(unzip('~/Downloads/birdclef-2023.zip', list = TRUE))
```

Unzipping.
```{r, eval=FALSE}
unzip('~/Downloads/birdclef-2023.zip',
      exdir = tempdir(),
      unzip = '/usr/bin/unzip')
```

### 2.3 Train
The .ogg files are converted using [`ffmpeg`](https://ffmpeg.org/), "[a] complete, cross-platform solution to record, convert and stream audio and video". The tagging procedure was performed by `voice::tag` using parallel processing. In order to save memory, the ogg to wav conversion, tagging and the wav files deletion are executed in the same routine. The processing took around 6 hours on a local machine with 12 CPUs, and the output is available in `birdclef2023::Esc`.

```{r, eval=FALSE}
# ogg files
oggFiles <- list.files(paste0(tempdir(), '/train_audio'), 
                       pattern = '.[Oo][Gg][Gg]$', 
                       full.names = TRUE, recursive = TRUE)
length(oggFiles)

# ogg directories
oggDirs <- unique(dirname(oggFiles))
length(oggDirs)

# wav files
new_pth <- paste0(tempdir(), '/wav/')
old_pth <- paste0(tempdir(), '/train_audio/')
wavFiles <- sub(old_pth, new_pth, oggFiles)
wavFiles <- sub('.ogg$', '.wav', wavFiles)

# wav directories
wavDirs <- unique(dirname(wavFiles))

# extended dataset
n <- length(oggDirs)
E_list <- vector('list', n)
names(E_list) <- basename(oggDirs)

# converting ogg to wav, tagging grouping by wav_path, deleting wav
for(i in 1:n){
  # garbage collector
  gc()
  
  # creating directories
  dir.create((wavDirs[i]), recursive = TRUE)
  
  # filtering by species_code
  oggTemp <- dir(oggDirs[i], full.names = TRUE)
  fltr <- grep(basename(oggDirs[i]), wavFiles)
  wavTemp <- wavFiles[fltr]
  
  # converting ogg to wav
  for(j in 1:length(oggTemp)){
    cmd <- paste0("ffmpeg -i ", oggTemp[j], " -ac 1 ", wavTemp[j])
    system(cmd)
  }
  
  # Extended dataset
  E <- dplyr::tibble(species_code = basename(oggDirs[i]),
                     wav_path = wavTemp)  
  
  # tagging
  E_list[[i]] <- voice::tag(E, groupBy = 'species_code',
                            features = c('f0', 'fmt', 'rf', 'rpf', 'rcf',
                                         'mfcc', 'zcr', 'rms', 'gain', 'rfc'))
  
  # binding
  E <- dplyr::bind_rows(E_list)
  write.csv(E, row.names = FALSE,
            file = gzfile(paste0(tempdir(), '/E_species_code.csv.gz')))
  
  # deleting wav files
  unlink(wavDirs[i], recursive = TRUE)
  
  # progress
  print(i/n)
}
```

### 2.4 Test

Converting the test file from ogg to wav format.
```{r, eval=FALSE}
# converting test data
oggFile <- paste0(tempdir(), '/test_soundscapes/soundscape_29201.ogg')
wavFile <- paste0(tempdir(), '/test_soundscapes/soundscape_29201.wav')
cmd <- paste0("ffmpeg -y -i ", oggFile, " -ac 1 ", wavFile)
system(cmd)
```

Splitting the 10 minutes file using `voice::splitw`. The procedure took around 25 seconds.
```{r, eval=FALSE}
# rttm
n <- 10*60/5
rttm <- dplyr::tibble(type = rep('SPEAKER', n),
                      file = rep(NA, n),
                      chnl = rep(1, n),
                      tbeg = seq(0,595,5),
                      tdur = rep(5,n),
                      ortho = rep(NA, n),
                      stype = rep(NA, n),
                      name = rep('A',n),
                      conf = rep(NA,n),
                      slat = rep(NA,n))
write.table(rttm, paste0(tempdir(), '/split5sec.rttm'),
            quote = FALSE, row.names = FALSE, col.names = FALSE)

# splitting
splitDir <- paste0(tempdir(), '/split')
dir.create(splitDir)
voice::splitw(wavFile, 
              fromRttm = paste0(tempdir(), '/split5sec.rttm'),
              toSplit = splitDir)

# renaming
fr <- dir(splitDir, full.names = TRUE)
row_id_temp <- paste0('soundscape_29201_', seq(5,600,5), '.wav')
to <- paste0(dirname(fr), '/', row_id_temp)
file.rename(fr,to)
```

Finally the tagging is applied to the test data. The procedure took less than 1 minute to run, and the output is available in `birdclef2023::Ets`.

```{r, eval=FALSE}
# Extended dataset
row_id_raw <- strsplit(row_id_temp, '.wav$')
row_id_raw <- sapply(row_id_raw, geti, 1)
E <- dplyr::tibble(row_id = row_id_raw,
                   wav_path = to)

# tagging grouping by row_id
tst <- voice::tag(E, groupBy = 'row_id',
                  features = c('f0', 'fmt', 'rf', 'rpf', 'rcf',
                               'mfcc', 'zcr', 'rms', 'gain', 'rfc'),
                  sortByGroupBy = FALSE,
                  mc.cores = parallel::detectCores())

# writing
write.csv(E, row.names = FALSE,
          file = gzfile(paste0(tempdir(), '/E_test_soundscapes.csv.gz')))
```


## 3. Modelling

### 3.1 Objects
The `birdclef2023::Esc` object contains 264 rows and 486 columns extracted via `voice::tag` from Kaggle BirdCLEF2023 `train_audio` directory.

```{r}
birdclef2023::Esc
```

The `birdclef2023::Ets` object contains 120 rows and 486 columns extracted via `voice::tag` from Kaggle BirdCLEF2023 `test_soundscapes` directory. Each row is associated with a 5 seconds section of `soundscape_29201.ogg` file.

```{r}
birdclef2023::Ets
```

### 3.2 Distances
The basic idea is to compare each row from `Ets` (test sample) with each row from `Esc` (reference values). The comparison is made through a metric derived from the Euclidean distance, taking the average of the valid values (not NA). We call `Average Length` and it is given by
\begin{equation}
AL=\sqrt{\frac{\sum_{i=1}^{n-n_{NA}} (x_i-y_i)^2}{n-n_{NA}}}
\end{equation} 

```{r}
# distances
D <- matrix(NA, nrow = nrow(birdclef2023::Ets), 
            ncol = nrow(birdclef2023::Esc))
rownames(D) <- birdclef2023::Ets$row_id
colnames(D) <- birdclef2023::Esc$species_code
```

The following code is completly not optimized, designed to be intuitive and run below the 2 hours limit defined in the Kaggle BirdCLEF2023 competition rules. Takes around 30 minutes to run on a local machine, and the output is available in `birdclef2023::D`.

```{r, eval=FALSE}
# calculating distances
k <- 0
for(i in 1:nrow(birdclef2023::Ets)){
  for(j in 1:nrow(birdclef2023::Esc)){
    k <- k+1
    d2 <- (birdclef2023::Ets[i,-1] - birdclef2023::Esc[j,-1])^2
    n <- length(d2)
    nNA <- sum(is.na(d2))
    D[i,j] <- sqrt(sum(d2, na.rm = TRUE)/(n-nNA))
    print(k/(nrow(birdclef2023::Ets)*nrow(birdclef2023::Esc)))
  }
}
```

```{r, eval=FALSE, echo=FALSE}
# save to rda file
save(D, file = './data/D.rda', compress = 'xz')
```

The probabilities are calculated considering the Gaussian function given by
\begin{equation}
g(x)=e^{-\left(\frac{x-\mu}{\sigma}\right)^2}
\end{equation}
$x$ represents the scaled distances, and the parameters used were $\mu=0$ and $\sigma=0.001$. The idea is to give more weight to distance values very close to zero, and quickly reduce the weight to distances that move away from zero, reason why the standard deviation is small (adjusted to 0.001).

```{r}
# probabilities matrix
P <- matrix(NA, nrow = nrow(birdclef2023::Ets), 
            ncol = nrow(birdclef2023::Esc))
rownames(P) <- birdclef2023::Ets$row_id
colnames(P) <- birdclef2023::Esc$species_code

# calculating probabilities
for(i in 1:nrow(P)){
  nm <- gaussian(scale(birdclef2023::D[i,]), 
                 mu = 0, sigma = 0.001)
  P[i,] <- nm/sum(nm)
}
```

After that, there is some cleaning and tidying up.

```{r}
block <- strsplit(rownames(P), '_')
block <- as.numeric(sapply(block, geti, 3))
submission <- tibble(block = block, 
                     row_id = rownames(P), 
                     as_tibble(P))

submission <- submission %>% 
  arrange(block) %>% 
  select(-block)

submission
```

Finally, the `submission.csv` file is written in the desired directory.

```{r, eval=FALSE}
write.csv(submission,
          file = paste0(tempdir(), '/submission.csv'))
```
